---
title: "Group Project: Research Question 1b"
author: "Saadana Balasubramanian"
date: "2026-07-26"
output: html_document

---
```{r chunk-setup,include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                       fig.align = "center")
```

# Research Question 1b

This section of the code addresses the second component of the first research question: "What is the prevalence of subjective sleep disturbance and what are its predictors in post liver transplant recipients?"

Each sleep quality score was modelled using demographic variables (Age,
Gender, BMI, time elapsed since transplant, liver diagnosis, recurrence
of disease, evidence of rejection/graft dysfunction, any fibrosis, renal
failure, depression, and corticosteroid use) as predictors.

## Chosen Models

The four instruments were recorded on different scales so two modeling
strategies were used depending on the instrument:

1.  **Continuous outcome — multiple linear regression.** PSQI, ESS, and
    AIS are all instruments that use numeric scores to indicate sleep
    quality. A multiple linear regression model was used to predict
    sleep quality while all demographic variables were entered as
    predictors. The output shows the estimated association of each
    predictor with sleep score while holding all other predictors
    constant.

2.  **Binary outcome — multiple logistic regression.** BSS uses a binary
    scale (1 = disturbance present). Additionally, PSQI, ESS, and AIS
    measurements were also converted to binary variables, using the
    clinically accepted thresholds to ensure the analysis is
    comprehensive. Multiple logistic regression models with the same set
    of predictors was used for these scales.

## Model Fitting and Comparison

There are 3 linear and 4 logistic models that were fitted on
`predictor_data`. This cleaned dataset contains all 11 predictors.

The coefficient (or odds ratio for logistic models), 95% confidence
interval, and p-value per predictor is derived from each model. From
this, a summary table of p-values across all 7 models is constructed to
compare which predictors are significant in each model.

------------------------------------------------------------------------

# R Code

## Load Packages & Import Data

```{r libraries}
# Packages used in this script -------------------------------------------
# car   : vif() for multicollinearity diagnostics
# knitr : kable() for formatted tables in the knitted report

library(car)
library(knitr)

#The existing cleaning pipeline is converted into an .R script 
knitr::purl("group_project.Rmd", output = "group_project_cleaning.R", quiet = TRUE)
source("group_project_cleaning.R")

# Confirm the objects we need now exist before continuing
stopifnot(exists("sleep_data"), exists("predictor_data"))
dim(predictor_data)
```

## Missing data check

Using lm() and glm() automatically drops missing values and conducts a
complete-case analysis by default. Before fitting the models, this
initial check is to find where missing data occurs. If there was a large
number of measured variables (ie. age, BMI, depression) that are missing
for certain outcomes then this would suggest those values are missing
not at random (NMAR). In this case, the sample isn't representative of
the population and complete-case analysis would be biased.

```{r missing-data-check}

# sapply() is used to apply the same function across every column in the list 
# this is used to count how many NA values are in each column

sapply(predictor_data[, c("PSQI", "ESS", "AIS", "PSQI_quality", "ESS_disturbance", "AIS_insomnia", "BSS", "age", "gender", "BMI", "transplant_time", "liver_diagnosis", "recurrence", "graft_dysfunction", "fibrosis", "renal_failure","depression", "corticosteroid")],
       function(x) sum(is.na(x)))

# Check is PSQI missing data > 30%
n_total <- nrow(predictor_data)

psqi_missing_pct <- sum(is.na(predictor_data$PSQI)) / n_total * 100
round(psqi_missing_pct, 1)

psqi_missing_pct

```

**Findings**: The table shows that there are no missing values for any
of the predictors which suggests that complete-case analysis is
appropriate to use. However, there are missing values for the outcome
variables (PSQI:80;ESS:15;AIS:5; BSS:6).

## Linear Regression Models (continuous variables)

All 11 predictors are entered simultaneously in each model. Each
coefficient represents the effect of that predictor while holding the
others constant.

```{r linear-models}
# lm() fits a multiple linear regression model. 

lm_PSQI <- lm(PSQI ~ age + gender + BMI + transplant_time + liver_diagnosis +
recurrence + graft_dysfunction + fibrosis + renal_failure + depression + 
corticosteroid, data = predictor_data)

lm_ESS <- lm(ESS ~ age + gender + BMI + transplant_time + liver_diagnosis +
recurrence + graft_dysfunction + fibrosis + renal_failure + depression + 
corticosteroid, data = predictor_data)

lm_AIS <- lm(AIS ~ age + gender + BMI + transplant_time + liver_diagnosis +
recurrence + graft_dysfunction + fibrosis + renal_failure + depression + 
corticosteroid, data = predictor_data)

# The estimated beta coefficients, p-value and standard error for each model
summary(lm_PSQI)
summary(lm_ESS)
summary(lm_AIS)

# 95% confidence intervals for the linear regression model coefficients, 
# rounded to 2 decimal places. 
round(confint(lm_PSQI), 2)
round(confint(lm_ESS), 2)
round(confint(lm_AIS), 2)

```

## Diagnostics and Multiple Comparisons

The assumptions of linear regression models are based on equal variance
of residuals, normality of residuals and linearity. These conditions
must be met to ensure that the model's predictions are accurate.
Diagnostic plots were used to evaluate if these assumptions have been
met. The models were also checked for multicollinearity using the
variance inflation factor (VIF). The multicollinearity check confirms
the assumption that none of the predictors in the models are correlated
with each other.

```{r linear-diagnostics}
# Residual diagnostic plots 
par(mfrow = c(2, 2))
plot(lm_PSQI, main = "PSQI") 
plot(lm_ESS,  main = "ESS")
plot(lm_AIS,  main = "AIS")
par(mfrow = c(1, 1))

# Multicollinearity check - VIF > 5 (or 10) flags problematic predictors
vif(lm_PSQI)
vif(lm_ESS)
vif(lm_AIS)
```

**Reasoning:** Using plot on a lm() creates 4 diagnostic plots.
par(mfrow = c(2, 2)) arranges these plots in a 2x2 grid. The first plot
checks linearity by ensuring that points are randomly scattered around
the horizontal line at 0. A curved pattern in this plot would indicate
that there is a non-linear relationship. The second qq plot checks
whether residuals are normally distributed by ensuring that points fall
along a diagonal line. Since the sample size is large (150-260 per
model), the data is expected to be normally distributed. The third plot
checks for equal variance of residuals by ensuring that points are
spread horizontally. The fourth plot checks if there are certain
observations that disproportionately affect the model.

**Findings:**

*Residual diagnostic plots:* There were no unusual findings in any of
the plots which confirms that the assumptions of the linear regression
model have been met (linearity, homoscedasticity, normality).

*Multicollinearity check:* There were no VIF scores above 5, which
confirms that none of the predictors in the model are correlated with
each other.

## Logistic Regression Models (Binary Sleep Outcomes)

All 11 predictors are entered simultaneously in each model. Each
coefficient represents the effect of that predictor while holding the
others constant.

```{r secondary-analysis}

# glm(..., family = binomial) fits a multiple logistic regression.
# Coefficients represent the log-odds of success
# exponentiate = TRUE converts log-odds to odds ratios

glm_PSQI_full <- glm(PSQI_quality ~ age + gender + BMI + transplant_time + 
liver_diagnosis + recurrence + graft_dysfunction + fibrosis + renal_failure +
depression + corticosteroid, data = predictor_data, family = binomial)

glm_ESS_full <- glm(ESS_disturbance ~ age + gender + BMI + transplant_time + liver_diagnosis + recurrence + graft_dysfunction + fibrosis + renal_failure +
depression + corticosteroid, data = predictor_data, family = binomial)

glm_AIS_full <- glm(AIS_insomnia ~ age + gender + BMI + transplant_time + 
liver_diagnosis + recurrence + graft_dysfunction + fibrosis + renal_failure +
depression + corticosteroid, data = predictor_data, family = binomial)

glm_BSS_full <- glm(BSS ~ age + gender + BMI + transplant_time + 
liver_diagnosis + recurrence + graft_dysfunction + fibrosis + renal_failure + 
depression + corticosteroid, data = predictor_data, family = binomial)

# The estimated beta coefficients, p-value and standard error for each model
summary(glm_PSQI_full)
summary(glm_ESS_full)
summary(glm_AIS_full)
summary(glm_BSS_full)

# Odds ratios

# For each logistic model, exponentiating the coefficients converts them from 
# log-odds to odds ratios. Values above 1 indicate higher odds of poor sleep 
# quality while values below 1 indicate lower odds. 
round(exp(glm_PSQI_full$coefficients), 2)
round(exp(glm_ESS_full$coefficients), 2)
round(exp(glm_AIS_full$coefficients), 2)
round(exp(glm_BSS_full$coefficients), 2)

# 95% confidence intervals for the odds ratios, rounded to 2 decimal places
round(exp(confint(glm_PSQI_full)), 2)
round(exp(confint(glm_ESS_full)), 2)
round(exp(confint(glm_AIS_full)), 2)
round(exp(confint(glm_BSS_full)), 2)

```

## Multicollinearity Check

```{r logistic-diagnostics}

# Multicollinearity check for the logistic models 
vif(glm_PSQI_full)
vif(glm_ESS_full)
vif(glm_AIS_full)
vif(glm_BSS_full)
```

**Findings:**

*Multicollinearity check:* There were no VIF scores above 5 for any of
the models, which confirms that none of the predictors in the model are
correlated with each other.

## Verifying Maximum Predictors

The maximum number of predictors for each model is identified using the
following formula: p =\< m/15. The following code identifies the number
of sample size in each linear regression model model, and the class with
the smallest sample size in each logistic regression model. Then, the
maximum number of predictors is found by dividing each sample size by 15
(the generally accepted rule). The final output is displayed as a
dataframe.

```{r verifying-max-predictors}

#nobs() finds the total number of observations that were used to predict the 
#output. This identifies the sample size of each linear regression model.
n_PSQI <- nobs(lm_PSQI)
n_ESS <- nobs(lm_ESS)
n_AIS <- nobs(lm_AIS)

#A frequency table is created which contains the number of observations within 
#each class of the outcome variable. The class with the smallest sample size 
#is used for this calculation. This is identified using 'min' and assigned to 
#a new variable. 

m_PSQI_glm <- min(table(predictor_data$PSQI_quality))
m_ESS_glm  <- min(table(predictor_data$ESS_disturbance))
m_AIS_glm  <- min(table(predictor_data$AIS_insomnia))
m_BSS_glm  <- min(table(predictor_data$BSS))

#A data frame is created to clearly display results. The first column includes 
#the model names. The second column lists the m value of each model and the last 
#column shows the maximum number of predictors used. 

data.frame(
  model = c("PSQI_lm", "ESS_lm", "AIS_lm",
            "PSQI_quality_glm", "ESS_disturbance_glm", "AIS_insomnia_glm", 
            "BSS_glm"),
  m = c(n_PSQI, n_ESS, n_AIS, m_PSQI_glm, m_ESS_glm, m_AIS_glm, m_BSS_glm),
  max_p = floor(c(n_PSQI, n_ESS, n_AIS, m_PSQI_glm, m_ESS_glm, m_AIS_glm, 
                  m_BSS_glm) / 15)
)
```

**Findings:** Based on this table, the current logistic regression
models exceed the maximum number of predictors that can be used (PSQI: 7
max predictors; ESS: 4 max predictors; AIS: 7 max predictors; BSS: 5 max
predictors). For these models, there is an increased risk of overfitting
and model inaccuracy.

## Secondary analysis with reduced models

Based on the results above, each logistic regression model above exceeds
the maximum number of predictors that can be included in each model. To
overcome this issue, reduced models were created by narrowing the number
of predictors used to avoid exceeding the maximum limit. These
predictors were chosen using a priori domain knowledge.

```{r logistic-models}

# glm(..., family = binomial) fits a multiple logistic regression.
# Coefficients represent the log-odds of success
# exponentiate = TRUE converts log-odds to odds ratios

# The outcome of each model has been encoded as binary based on the clinically verified thresholds during the data processing procedure. 

# PSQI logistic regression with 7 df
glm_PSQI_reduced <- glm(PSQI_quality ~ depression + corticosteroid + BMI +
                          transplant_time + renal_failure + age + gender,
                        family = binomial, data = predictor_data)

# ESS logistic regression with 4 df
glm_ESS_reduced <- glm(ESS_disturbance ~ depression + corticosteroid + BMI +
                         transplant_time,
                       family = binomial, data = predictor_data)

# AIS logistic regression with 7 df
glm_AIS_reduced <- glm(AIS_insomnia ~ depression + corticosteroid + BMI +
                         transplant_time + renal_failure + age + gender,
                       family = binomial, data = predictor_data)

# BSS logistic regression with 5 df 
glm_BSS_reduced <- glm(BSS ~ depression + corticosteroid + BMI +
                         transplant_time + age,
                       family = binomial, data = predictor_data)

# The estimated beta coefficients, p-value and standard error for each model
summary(glm_PSQI_reduced)
summary(glm_ESS_reduced)
summary(glm_AIS_reduced)
summary(glm_BSS_reduced)

# Odds ratios
# For each logistic model, exponentiating the coefficients converts them from 
# log-odds to odds ratios. Values above 1 indicate higher odds of poor sleep 
# quality while values below 1 indicate lower odds. 

round(exp(glm_PSQI_reduced$coefficients), 2)
round(exp(glm_ESS_reduced$coefficients), 2)
round(exp(glm_AIS_reduced$coefficients), 2)
round(exp(glm_BSS_reduced$coefficients), 2)

# 95% confidence intervals for the odds ratios, rounded to 2 decimal places
round(exp(confint(glm_PSQI_reduced)), 2)
round(exp(confint(glm_ESS_reduced)), 2)
round(exp(confint(glm_AIS_reduced)), 2)
round(exp(confint(glm_BSS_reduced)), 2)

```

# Comparing Full vs Reduced Logistic Models

```{r compare-logistic-models}

# Likelihood ratio test (LRT) is used to compare full vs reduced logistic models
# anova() with test = "Chisq" compares the change in deviance between two 
# nested models. The null hypothesis is that the reduced model fits the data
# well while the alternative hypothesis is that the reduced model does fit the 
# data better than the full model. A significant p-value means the full model 
# fits significantly better. A non-significant p-value supports using the 
# reduced model.

anova(glm_PSQI_reduced, glm_PSQI_full, test = "Chisq")
anova(glm_ESS_reduced,  glm_ESS_full,  test = "Chisq")
anova(glm_AIS_reduced,  glm_AIS_full,  test = "Chisq")
anova(glm_BSS_reduced,  glm_BSS_full,  test = "Chisq")
```

**Findings:** The reduced model of PSQI, AIS and BSS fits the data
better. The full model of ESS fits the data significantly better.

# Comparing Models & Predictors

## Summary Table 1: Reduced model comparison across various sleep quality measurements

A summary table is created to compare each sleep quality measurement and
the set of predictors that are significantly associated with the outcome
in each. The reduced logistic regression models are used in this
comparison. The p-value of each model's coefficient matrix is extracted
from the summary() output to create this table. For predictors that are
not present in the reduced model, NA is used in their place to indicate
they were removed from the model.

```{r comparision-table-1}

# The 4th column of summary(model)$coefficients has the p-values for
# lm() and glm(). Row 1 (containing the intercept) is dropped since we are
# only comparing each predictor.

p_PSQI     <- summary(lm_PSQI)$coefficients[-1, 4]
p_ESS      <- summary(lm_ESS)$coefficients[-1, 4]
p_AIS      <- summary(lm_AIS)$coefficients[-1, 4]
p_PSQI_reduced <- summary(glm_PSQI_reduced)$coefficients[-1, 4]
p_ESS_reduced  <- summary(glm_ESS_reduced)$coefficients[-1, 4]
p_AIS_reduced  <- summary(glm_AIS_reduced)$coefficients[-1, 4]
p_BSS_reduced      <- summary(glm_BSS_reduced)$coefficients[-1, 4]


# all_predictor is a new variable containing all of the variable names included in the models 
all_predictors <- unique(c(names(p_PSQI), names(p_ESS), names(p_AIS),
                            names(p_PSQI_reduced), names(p_ESS_reduced),
                            names(p_AIS_reduced), names(p_BSS_reduced)))

# Unlike the full models, the reduced models don't all use the same predictors. Because of this, using cbind() would cause an error or mislabel columns. For example, there's no position 5 in p_ESS_bin because a 5th variable wasn't included. In this case, another variable would be substituted in its place and would risk mislabelling.

# Using match() aligns each model's p-values to its corresponding predictor, using a common set of predictor names. If that predictor wasn't included in the model, NA fills that spot. 

pval_matrix_1 <- cbind(
  PSQI_lm              = p_PSQI[match(all_predictors, names(p_PSQI))],
  ESS_lm               = p_ESS[match(all_predictors, names(p_ESS))],
  AIS_lm               = p_AIS[match(all_predictors, names(p_AIS))],
  PSQI_quality_glm     = p_PSQI_reduced[match(all_predictors, names(p_PSQI_reduced))],
  ESS_disturbance_glm  = p_ESS_reduced[match(all_predictors, names(p_ESS_reduced))],
  AIS_insomnia_glm     = p_AIS_reduced[match(all_predictors, names(p_AIS_reduced))],
  BSS_glm              = p_BSS_reduced[match(all_predictors, names(p_BSS_reduced))]
)

#Row names of the matrix are assigned using the common set of variable names
rownames(pval_matrix_1) <- all_predictors

round(pval_matrix_1, 3)
kable(round(pval_matrix_1, 3), caption = "P-values for each reduced model's predictors across the seven fitted models (reduced models)")
```

## Summary Table 2: Full model comparision across various sleep quality measurements

A summary table is created to compare each sleep quality measurement and
the full set of predictors. The p-value of each model's coefficient
matrix is extracted from the summary() output to create this table. This
full model table is included as a part of the secondary analysis to
report all tested predictors with full transparency, ensuring the
analysis comprehensive. The current results can be compared with the
reduced table to confirm that the excluded predictors were not
significant in the model and to compare whether these conclusions are
robust.

```{r comparison-table-2}

# The 4th column of summary(model)$coefficients has the p-values for
# lm() and glm(). Row 1 (containing the intercept) is dropped since we are
# only comparing each predictor.

p_PSQI <- summary(lm_PSQI)$coefficients[-1, 4]
p_ESS  <- summary(lm_ESS)$coefficients[-1, 4]
p_AIS  <- summary(lm_AIS)$coefficients[-1, 4]
p_PSQI_full <- summary(glm_PSQI_full)$coefficients[-1, 4]
p_ESS_full  <- summary(glm_ESS_full)$coefficients[-1, 4]
p_AIS_full  <- summary(glm_AIS_full)$coefficients[-1, 4]
p_BSS_full     <- summary(glm_BSS_full)$coefficients[-1, 4]

# All 7 models use the same 11 predictors and the coefficient names 
# are the same in every model. cbind() combines the columns with the p-value
# for each matrix. Each row represents the predictor and each column 
# represents the model that was used.
pval_matrix_2 <- cbind(
  PSQI_lm = p_PSQI,
  ESS_lm = p_ESS,
  AIS_lm = p_AIS,
  PSQI_quality_glm = p_PSQI_full,
  ESS_disturbance_glm = p_ESS_full,
  AIS_insomnia_glm = p_AIS_full,
  BSS_glm = p_BSS_full
)

round(pval_matrix_2, 3)
kable(round(pval_matrix_2, 3), caption = "P-values for each full model's predictors across the seven fitted models (complete models)")
```

## Significance flag table

A summary table is created with asterisks flagging predictors that are
significantly associated (p \< 0.05) with sleep quality for quick visual
scanning.

```{r significance-flags}
sig_matrix <- ifelse(pval_matrix_1 < 0.05, "*", "")
kable(sig_matrix, caption = "Significance flags (p < 0.05) by predictor and model")
```

## Consistently significant predictors across models

The following table shows the number of models where each predictor is
significant in comparison to the total number of models that were
tested. This identifies predictors that are more consistently associated
with sleep quality across different sleep quality measurement scales.

```{r consistency-summary}

# apply(..., 1, sum) goes row by row (i.e. predictor by predictor) and
# counts how many of the 7 models have a significant p-value (p < 0.05) for that 
# row.
n_sig_models <- apply(pval_matrix_1 < 0.05, 1, sum, na.rm = TRUE)
n_tested_models <- apply(!is.na(pval_matrix_1), 1, sum)

consistency_summary <- data.frame(
  Predictor          = rownames(pval_matrix_1),
  Models_significant = n_sig_models,
  Models_tested       = n_tested_models
)

# This sorts the table so the most consistently significant predictors appear first.
consistency_summary <- consistency_summary[order(-consistency_summary$Models_significant), ]

kable(consistency_summary, row.names = FALSE,
      caption = "Number of models in which each predictor is significant (p < 0.05), 
      out of models where it was tested")
```
