---
title: "Group Project"
author: "Ranmali Attapattu"
date: "2026-07-25"
output: html_document
---


# Initial Data Exploration and Cleaning

This section of the code focuses on the initial cleaning & familiarization of the data for later analysis.
This analysis uses the following packages: `dplyr`, `ggplot`, `knitr`, `tidyr`

```{r load-data, results='hide', message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)

project_data <- read.csv("project_data.csv")
```

## Initial Exploration and Cleaning
### Initial Check
```{r initial-check, results='hide'}
head(project_data)
dim(project_data)
str(project_data)
summary(project_data)
sum(duplicated(project_data))

```
An initial exploration of the dataset showed that it contains **268 observations** and a mix of *demographic, clinical, medication,* and *patient-reported* outcome variables. No duplicate observations were identified. Most clinical variables were complete, while BMI (23 missing values) and several questionnaire-based measures exhibited missingness. In particular, the Pittsburgh Sleep Quality Index (85 missing values), Epworth Sleepiness Scale (17 missing values before cleaning), Athens Insomnia Scale (6), Berlin Sleepiness Scale (6), and SF-36 quality-of-life measures (21 each) contained incomplete observations. Additionally, **six variables (X.1–X.6)** contained *only missing values* and appear to be empty columns and were removed from the analysis subset as they contained no information.

Range checks were performed to identify potential data entry errors or implausible values. Demographic variables, clinical characteristics, sleep measures, and quality of life scores were within expected ranges. One observation had an **Epworth Sleepiness Scale (ESS)** value of 26, which exceeded the commonly accepted scoring range of 0–24. Because this value could not be verified and was outside the valid range, it was treated as missing for subsequent analyses.
```{r ess-correction, results='hide'}

project_data <- project_data %>%
  mutate(
    Epworth.Sleepiness.Scale = ifelse(Epworth.Sleepiness.Scale > 24, NA, Epworth.Sleepiness.Scale)
  )

```
Summary statistics indicate that participants ranged in age from 18 to 80 years (mean = 52.78 years), with an average BMI of 27.02 kg/m² among those with recorded values. The average time since transplant was 6.49 years, ranging from 0.08 to 23.02 years. Overall, the dataset appears well suited for analysis, although variables with substantial missing data will require careful handling before conducting statistical analyses.

## Cleaning
Given the large number of variables in the original dataset, only variables relevant to the research questions were retained. Variable labels and coding were recoded according to the data dictionary.
```{r subsetting, echo=FALSE}
sleep_data <- project_data %>%
  select(
    Age,
    Gender,
    BMI,
    Time.from.transplant,
    Liver.Diagnosis,
    Recurrence.of.disease,
    Rejection.graft.dysfunction,
    Any.fibrosis,
    Renal.Failure,
    Depression,
    Corticoid,
    Epworth.Sleepiness.Scale,
    Excessive.daytime.sleepiness,
    Pittsburgh.Sleep.Quality.Index.Score,
    Poor.sleep.quality,
    Berlin.Sleepiness.Scale,
    Insomnia,
    Athens.Insomnia.Scale,
    SF36.PCS,
    SF36.MCS
  )

sleep_data <- sleep_data %>%
  rename(
    age = Age,
    gender = Gender,
    transplant_time = Time.from.transplant,
    liver_diagnosis = Liver.Diagnosis,
    recurrence = Recurrence.of.disease,
    graft_dysfunction = Rejection.graft.dysfunction,
    fibrosis = Any.fibrosis,
    renal_failure = Renal.Failure,
    depression = Depression,
    corticosteroid = Corticoid,
    ESS = Epworth.Sleepiness.Scale,
    ESS_disturbance = Excessive.daytime.sleepiness,
    PSQI = Pittsburgh.Sleep.Quality.Index.Score,
    PSQI_quality = Poor.sleep.quality,
    BSS = Berlin.Sleepiness.Scale,
    AIS = Athens.Insomnia.Scale,
    AIS_insomnia = Insomnia,
    SF36_PCS = SF36.PCS,
    SF36_MCS = SF36.MCS
  )
```


### Recoding Demographic and Clinical Variables
```{r recoding}
sleep_data$gender = factor(sleep_data$gender,
                levels = c(1, 2),
                labels = c("Male", "Female"))

sleep_data$liver_diagnosis = factor(sleep_data$liver_diagnosis,
                         levels = c(1, 2, 3, 4, 5),
                         labels = c("Hep C", "Hep B", "PSC/PBC/AHA", "alcohol", "other"))

sleep_data$recurrence = factor(sleep_data$recurrence,
                    levels = c(1, 0),
                    labels = c("Yes", "No"))

sleep_data$graft_dysfunction = factor(sleep_data$graft_dysfunction,
                           levels = c(1, 0),
                           labels = c("Yes", "No"))
                         
sleep_data$fibrosis = factor(sleep_data$fibrosis,
                  levels = c(1, 0),
                  labels = c("Yes", "No"))

sleep_data$renal_failure = factor(sleep_data$renal_failure,
                       levels = c(1, 0),
                       labels = c("Yes", "No"))

sleep_data$depression = factor(sleep_data$depression,
                    levels = c(1, 0),
                    labels = c("Yes", "No"))

sleep_data$corticosteroid = factor(sleep_data$corticosteroid,
                        levels = c(1, 0),
                        labels = c("Yes", "No"))

```

After re-coding, the demographic data is easier to interpret.

```{r table1, results='asis', echo=FALSE, fig.align='centre'}
kable(sleep_data[1:5,], caption = "First five observations of the analysis subset, with variables renamed for clarity.")
```

### Recoding Sleep Disturbance Variables
In order to accommodate multiple analysis angles, sleep disturbance was stored using both the **continuous** questionnaire scores and the clinically meaningful **binary** classifications. Continuous scores were retained to assess the severity of sleep symptoms and their relationship with health-related quality of life. Binary sleep disturbance variables were retained and relabeled using descriptive factor levels to facilitate interpretation and support prevalence estimation and logistic regression analyses.
```{r binary-vars, echo=FALSE}

sleep_data <- sleep_data %>%
  mutate(
    PSQI_quality = factor(PSQI_quality,
                              levels = c(0, 1),
                              labels = c("Adequate quality", "Poor sleep quality")
                              ),
    ESS_disturbance = factor(ESS_disturbance,
                             levels = c(0, 1),
                             labels = c("No disturbance", "Disturbance")
                             ),
    AIS_insomnia = factor(AIS_insomnia,
                          levels = c(0, 1),
                          labels = c("No insomnia", "Insomnia")
                          ),
    BSS = factor(BSS,
                 levels = c(0, 1),
                 labels = c("No disturbance", "Disturbance")
                 )
    )

binary_vars <- sleep_data[, c("PSQI_quality", "ESS_disturbance", "AIS_insomnia", "BSS")]

kable(binary_vars[1:5,], caption = "Binary sleep disturbance variables.")

```

###  Missing Values
Several variables had minimal missingness (<10%), while questionnaire-based measures showed higher levels of missing data, particularly PSQI (31.7%). Since removing all observations with missing values would substantially reduce the sample size and potentially introduce bias, analyses were performed using available cases for each outcome measure.
```{r missing_values, echo=FALSE, results='asis'}
kable(colSums(is.na(sleep_data)), col.names = c("Variable", "Missing Value Count"), caption = "A summary of all missing values in the analysis subset.")
```
Although regression models require complete datasets, retaining all available observations maximizes statistical power and avoids unnecessary loss of information. Regression analyses involving predictor variables requiring complete data were conducted using the `predictor_data` subset, while descriptive analyses used all available observations in the `sleep_data` subset.
```{r predictor_data}
predictor_data <- sleep_data %>%
  drop_na(age, gender, BMI, transplant_time, 
          liver_diagnosis, recurrence,
          graft_dysfunction, fibrosis,
          renal_failure, depression,
          corticosteroid)
```
